---
title: "Static site occupancy model - Unmarked"
author: "Markus Fjellstad Israelsen"
date: "25 januar 2018"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Import detection/nondetection data and divide into year (to do single-season occupancy modelling on each year separately).
Import matrix of sites and their latitude + longitude coordinates.
Import line observations
Import masl, habitat and mean june temperatures for each site
```{r}
library(readr)
yfjellrype <- read_delim("C:/Users/marku/Desktop/Master/Dataset/mydata.csv", ";", escape_double = FALSE, trim_ws = TRUE)
View(yfjellrype)

# Split yfjellrype into each year
y2011 = yfjellrype[ , 1:20]
y2012 = yfjellrype[, 21:40]
y2013 = yfjellrype[, 41:60]
y2014 = yfjellrype[, 61:80]
y2015 = yfjellrype[, 81:100]
y2016 = yfjellrype[, 101:120]

library(readr)
First_fjellrype_lat_long = read_csv("C:/Users/marku/Desktop/Master/Dataset/First fjellrype lat long.csv")
View(First_fjellrype_lat_long)

library(readr)
Line_observations <- read_csv("C:/Users/marku/Desktop/Master/Dataset/Line observations.csv")
View(Line_observations)

library(readr)
Covariater_Midpunkt_UTMSone33 <- read_delim("C:/Users/marku/Desktop/Master/Dataset/Covariater_Midpunkt_UTMSone33.csv", 
    ";", escape_double = FALSE, trim_ws = TRUE)
View(Covariater_Midpunkt_UTMSone33)
Covariater_Midpunkt_UTMSone33 = Covariater_Midpunkt_UTMSone33[- c(494), ]
```


Load unmarked package
```{r}
library(unmarked)
```


Start with creating a vector containing the sites that are known to be occupied (for the "knownOcc" argument in the single-season model). First, the variable sitenum is created to be able to identify which rownumber corresponds to which sitenumber (for instance, row [45] --> sitenumber 432)
```{r}
sitenum = c()
sitenum = as.matrix(First_fjellrype_lat_long$RUTENR, ncol = 1)
sitenum = matrix(sitenum, ncol = 1, dimnames = list(c(), c("site")))
head(sitenum)

```


Identifying the rows for the known occupancy of the 2011 detection/nondetection data
```{r}
lineobs = data.frame(Line_observations)
lineobs$Art = NULL
line2011 = c()

# for loop that appends the rows from the line observations with the year 2011 to the variable line2011
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2011) {
    line2011 = append(line2011, lineobs[row, 1])
  }
}
line2011 = matrix(line2011, ncol = 1)
line2011 = cbind(line2011, rep(2011, length(line2011)))
line2011 = matrix(line2011, ncol = 2, dimnames = list(c(), c("sitenumber", "year")))
line2011 = data.frame(line2011)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2011 = line2011$sitenumber
count = 0
row2011 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2011)){
    row2011 = append(row2011, count) # row2011 is the variable with the rownumber for each site known to be occupied
  }
}

```

Create variables for the unstandardized, original covariates 
```{r}
sitenr = Covariater_Midpunkt_UTMSone33$RuteNr
masl.orig = Covariater_Midpunkt_UTMSone33$H?ydeOverHavet
habitat.orig = Covariater_Midpunkt_UTMSone33$Norut_Habitat1_2
precip.year.orig = Covariater_Midpunkt_UTMSone33$?rsnedb?r
avg.temp.june.orig = Covariater_Midpunkt_UTMSone33$Juni_Middeltemp
latitude.orig = First_fjellrype_lat_long$Latitude
longitude.orig = First_fjellrype_lat_long$Longitude

```

Overview of covariates
```{r}
covs = cbind(masl.orig, precip.year.orig, avg.temp.june.orig, latitude.orig, longitude.orig)
par(mfrow = c(3,3))
for(i in 1:5){
hist(covs[,i], breaks = 50, col = "grey", main = colnames(covs)[i])
}
pairs(cbind(masl.orig, precip.year.orig, avg.temp.june.orig, latitude.orig, longitude.orig))
```


Standardize the original covariates (Standardizing to get the variables on the same scale?)
```{r}
mean.masl = mean(masl.orig, na.rm = TRUE)
mean.precip.year.orig = mean(precip.year.orig, na.rm = TRUE)
mean.avg.temp.june.orig = mean(avg.temp.june.orig, na.rm = TRUE)
mean.latitude.orig = mean(latitude.orig, na.rm= TRUE) # Is it ok to standardize latitude and longitude? 
mean.longitude.orig = mean(longitude.orig, na.rm = TRUE) 

sd.masl = sd(masl.orig, na.rm = TRUE)
sd.precip.year.orig = sd(precip.year.orig, na.rm = TRUE)
sd.avg.temp.june.orig = sd(avg.temp.june.orig, na.rm = TRUE)
sd.latitude.orig = sd(latitude.orig, na.rm = TRUE)
sd.longitude.orig = sd(longitude.orig, na.rm = TRUE)

masl = (masl.orig - mean.masl) / sd.masl
precip.year = (precip.year.orig - mean.precip.year.orig) / sd.precip.year.orig
avg.temp.june = (avg.temp.june.orig - mean.avg.temp.june.orig) / sd.avg.temp.june.orig
latitude = (latitude.orig - mean.latitude.orig) / sd.latitude.orig
longitude = (longitude.orig - mean.longitude.orig) / sd.longitude.orig

```


Add detection/nondetection data and covariates together in a matrix, which makes it possible to remove the rows where there are missing covariate information. 
Overwrite the covariate variables based on the new dataset.
THE OCCU FUNCTION ONLY REMOVES ROWS WITH ALL NA's
```{r}
y2011.cov = cbind(y2011, masl, precip.year, avg.temp.june, latitude, longitude, habitat.orig, sitenr)
View(y2011.cov)
y2011.cov = cbind(y2011, masl.orig, precip.year.orig, avg.temp.june.orig, latitude.orig, longitude.orig, habitat.orig, sitenr)

y2011.cov = y2011.cov[complete.cases(y2011.cov[ , 21:25]),] # Removing rows where either of the covariates have NAs (including the y values)

y2011.cov = y2011.cov[- c(480),] # Manually removed last row because the habitat.orig had a NA value in the last row that was not caught by the complete.cases function

masl.2011 = y2011.cov$masl
precip.2011 = y2011.cov$precip.year
avg.temp.june.2011 = y2011.cov$avg.temp.june
latitude.2011 = y2011.cov$latitude
longitude.2011 = y2011.cov$longitude
habitat.2011 = y2011.cov$habitat.orig
latitude.2011 = y2011.cov$latitude
sitenr.2011 = y2011.cov$sitenr

masl.2011 = y2011.cov$masl.orig
precip.2011 = y2011.cov$precip.year.orig
avg.temp.june.2011 = y2011.cov$avg.temp.june.orig
latitude.2011 = y2011.cov$latitude.orig
longitude.2011 = y2011.cov$longitude.orig
habitat.2011 = y2011.cov$habitat.orig
latitude.2011 = y2011.cov$latitude.orig
sitenr.2011 = y2011.cov$sitenr



```

#### UnmarkedFrameOccu - 2011
Create unmarkedFrameOccu by separating the sitecovariates from the observational covariates
```{r}
# Create the covariates for 'obsCovs', as they need to be in a matrix form
obs.habitat.2011 = matrix(data = rep(habitat.2011, 20), nrow = 479, ncol = 20)
obs.avg.temp.june.2011 = matrix(data = rep(avg.temp.june.2011, 20), nrow = 479, ncol = 20)
obs.masl.2011 = matrix(data = rep(masl.2011, 20), nrow = 479, ncol = 20)

# UnmarkedFrameOccu
umf = unmarkedFrameOccu(y = y2011.cov[1:20], siteCovs = data.frame(masl.2011, avg.temp.june.2011, habitat.2011, latitude.2011), obsCovs = list(obs.masl.2011 = obs.masl.2011, obs.avg.temp.june.2011 = obs.avg.temp.june.2011, obs.habitat.2011 = obs.habitat.2011))

```


#### Model selection on the detection part - 2011. 
```{r}
# formula = ~ detection ~ occupancy

# Effect of habitat on detection prob
dmod1 = occu(formula = ~ 1 ~ 1, data = umf, knownOcc = row2011)
summary(dmod1)
dmod2 = occu(formula = ~ obs.habitat.2011 ~ 1, data = umf, knownOcc = row2011)
summary(dmod2)
# Insert the fitted models into a "fitList" and rank them by AIC
fms = fitList("p(.)psi(.)" = dmod1,
              "p(habitat)psi(.)" = dmod2)
(dmodel.sel = modSel(fms))
# NaNs produced! Do not think it worked the way you thought it would.

# Effect of mean june temp on detection prob
dmod3 = occu(formula = ~ obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod4 = occu(formula = ~ obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod5 = occu(formula = ~ obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + I(obs.avg.temp.june.2011^3) ~ 1, data = umf, knownOcc = row2011)

# Insert the fitted models into a "fitList" and rank them by AIC
dms1 = fitList("p(.)psi(.)" = dmod1,
               "p(temp)psi(.)" = dmod3,
               "p(temp + temp2)psi(.)" = dmod4,
               "p(temp + temp2 + temp3)psi(.)" = dmod5)
(dmodel.sel1 = modSel(dms1))
# Model 4 best, but only with AICdelta of 0.27 to model 3 which has 1 less parameter and is the one I will choose? 

# Effect of mean june temp on detection prob
dmod6 = occu(formula = ~ obs.masl.2011 ~ 1, data = umf, knownOcc = row2011)

dmod7 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod8 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + I(obs.masl.2011^3) ~ 1, data = umf, knownOcc = row2011)

# Insert the fitted models into a "fitList" and rank them by AIC
dms2 = fitList("p(.)psi(.)" = dmod1,
               "p(elev)psi(.)" = dmod6,
               "p(elev + elev2)psi(.)" = dmod7,
               "p(elev + elev2 + elev3)psi(.)" = dmod8)
(dmodel.sel2 = modSel(dms2))

# Effect of elevation and mean june temp on detection prob
dmod9 = occu(formula = ~ 1 ~ 1, data = umf, knownOcc = row2011) 

dmod10 = occu(formula = ~ obs.masl.2011 ~ 1, data = umf, knownOcc = row2011)

dmod11 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod12 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + I(obs.masl.2011^3) ~ 1, data = umf, knownOcc = row2011)

dmod13 = occu(formula = ~ obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod14 = occu(formula = ~ obs.masl.2011 + obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod15 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod16 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + I(obs.masl.2011^3) + obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod17 = occu(formula = ~ obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod18 = occu(formula = ~ obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + I(obs.avg.temp.june.2011^3) ~ 1, data = umf, knownOcc = row2011)

dmod19 = occu(formula = ~ obs.masl.2011 + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod20 = occu(formula = ~ obs.masl.2011 + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + I(obs.avg.temp.june.2011^3) ~ 1, data = umf, knownOcc = row2011)

dmod21 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod22 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + I(obs.masl.2011^3) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod23 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + I(obs.avg.temp.june.2011^3) ~ 1, data = umf, knownOcc = row2011)

dmod24 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + I(obs.masl.2011^3) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + I(obs.avg.temp.june.2011^3) ~ 1, data = umf, knownOcc = row2011)

dmod25 = occu(formula = ~ obs.masl.2011 + obs.avg.temp.june.2011 + obs.masl.2011:obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod26 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod27 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dmod28 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ 1, data = umf, knownOcc = row2011)


# Insert the fitted models into a "fitList" and rank them by AIC
dms3 = fitList("p(.)psi(.)" = dmod9,
               "p(elev)psi(.)" = dmod10,
               "p(elev + elev2)psi(.)" = dmod11,
               "p(elev + elev2 + elev3)psi(.)" = dmod12,
               "p(temp)psi(.)" = dmod13,
               "p(elev + temp)psi(.)" = dmod14,
               "p(elev + elev2 + temp)psi(.)" = dmod15,
               "p(elev + elev2 + elev3 + temp)psi(.)" = dmod16,
               "p(temp + temp2)psi(.)" = dmod17,
               "p(temp + temp2 + temp3)psi(.)" = dmod18,
               "p(elev + temp + temp2)psi(.)" = dmod19,
               "p(elev + temp + temp2 + temp3)psi(.)" = dmod20,
               "p(elev + elev2 + temp + temp2)psi(.)" = dmod21,
               "p(elev + elev2 + elev3 + temp + temp2)psi(.)" = dmod22,
               "p(elev + elev2 + temp + temp2 + temp3)psi(.)" = dmod23,
               "p(elev + elev2 + elev3 + temp + temp2 + temp3)psi(.)" = dmod24,
               "p(elev + temp + elev:temp)psi(.)" = dmod25,
               "p(elev + elev2 + temp + temp2 + elev:temp)psi(.)" = dmod26,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp)psi(.)" = dmod27,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(.)" = dmod28)
(dmodel.sel3 = modSel(dms3))

# Effect of latitude on detection probability
dmod29 = occu(formula = ~ latitude.2011 ~ 1, data = umf, knownOcc = row2011)

dmod30 = occu(formula = ~ latitude.2011 + I(latitude.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod31 = occu(formula = ~ latitude.2011 + I(latitude.2011^2) + I(latitude.2011^3) ~ 1, data = umf, knownOcc = row2011)

dms4 = fitList("p(.)psi(.)" = dmod9,
               "p(latitude)psi(.)" = dmod29,
               "p(latitude + latitude2)psi(.)" = dmod30,
               "p(latitude + latitude2 + latitude3)psi(.)" = dmod31)
(dmodel.sel4 = modSel(dms4))

# Effect of elevation, temp and lat
dmod32 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) + latitude.2011 ~ 1, data = umf, knownOcc = row2011)

dmod33 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) + latitude.2011 + I(latitude.2011^2) ~ 1, data = umf, knownOcc = row2011)

dmod34 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) + latitude.2011 + I(latitude.2011^2) + latitude.2011:obs.avg.temp.june.2011 ~ 1, data = umf, knownOcc = row2011)

dms5 = fitList("p(.)psi(.)" = dmod9, 
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(.)" = dmod28,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2 + latitude)psi(.)" = dmod32,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2 + latitude + latitude2)psi(.)" = dmod33,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2 + latitude + latitude2 + latitude:temp)psi(.)" = dmod34)
(dmodel.sel5 = modSel(dms5))

```


#### Model selection on the occupancy part - 2011
The correlation between 'masl.2011' and 'avg.temp.june.2011' = -0.4461618, and this is quite high, and could therefore affect the results I'm getting?
```{r}
# formula = ~ detection ~ occupancy
# Looking for the effects of elevation first
om1 = occu(formula = ~ 1 ~ 1, data = umf, knownOcc = row2011)

om2 = occu(formula = ~ 1 ~ masl.2011, data = umf, knownOcc = row2011)

om3 = occu(formula = ~ 1 ~ masl.2011 + I(masl.2011^2), data = umf, knownOcc = row2011)

om4 = occu(formula = ~ 1 ~ masl.2011 + I(masl.2011^2) + I(masl.2011^3), data = umf, knownOcc = row2011)

# Insert the fitted models into a "fitList" and rank them by AIC
oms1 = fitList("p(.)psi(.)" = om1,
              "p(.)psi(elevation)" = om2,
              "p(.)psi(elevation + elevation2)" = om3,
              "p(.)psi(elevation + elevation2 + elevation3)" = om4)
(omodel.sel1 = modSel(oms1))

# Look for the effects of mean june temperature
om5 = occu(formula = ~ 1 ~ avg.temp.june.2011, data = umf, knownOcc = row2011)

om6 = occu(formula = ~ 1 ~ avg.temp.june.2011 + I(avg.temp.june.2011^2), data = umf, knownOcc = row2011)

om7 = occu(formula = ~ 1 ~ avg.temp.june.2011 + I(avg.temp.june.2011^2) + I(avg.temp.june.2011^3), data = umf, knownOcc = row2011) 

# Insert the fitted models into a "fitList" and rank them by AIC
oms2 = fitList("p(.)psi(.)" = om1,
              "p(.)psi(temp)" = om5,
              "p(.)psi(temp + temp2)" = om6,
              "p(.)psi(temp + temp2 + temp3)" = om7)

(omodel.sel2 = modSel(oms2))

# Look for the effects of habitat
om8 = occu(formula = ~ 1 ~ habitat.2011, data = umf, knownOcc = row2011)

# Insert the fitted models into a "fitList" and rank them by AIC
oms3 = fitList("p(.)psi(.)" = om1,
              "p(.)psi(habitat)" = om8)

(omodel.sel3 = modSel(oms3))

# Look for the effects of elevation when elevation and temp are in the detection prob.
om9 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ masl.2011, data = umf, knownOcc = row2011)

om10 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ masl.2011 + I(masl.2011^2), data = umf, knownOcc = row2011)

om11 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ masl.2011 + I(masl.2011^2) + I(masl.2011^3), data = umf, knownOcc = row2011)

oms4 = fitList("p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(.)" = dmod28,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(elev)" = om9,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(elev + elev2)" = om10,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(elev + elev2 + elev3)" = om11)
(omodel.sel4 = modSel(oms4))

# Look for the effects of temperature when elevation and temp are in the detection prob.
om12 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ avg.temp.june.2011, data = umf, knownOcc = row2011)

om13 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ avg.temp.june.2011 + I(avg.temp.june.2011^2), data = umf, knownOcc = row2011)

om14 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ avg.temp.june.2011 + I(avg.temp.june.2011^2) + I(avg.temp.june.2011^3), data = umf, knownOcc = row2011)

oms5 = fitList("p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(.)" = dmod28,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(temp)" = om12,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(temp + temp2)" = om13,
               "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(temp + temp2 + temp3)" = om14)
(omodel.sel5 = modSel(oms5))

# Look for the effects of habitat when elevation and temp are in the detection prob.
om15 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ habitat.2011, data = umf, knownOcc = row2011)

# Insert the fitted models into a "fitList" and rank them by AIC
oms6 = fitList("p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(.)" = dmod28,
              "p(elev + elev2 + temp + temp2 + elev:temp + elev2:temp + elev:temp2)psi(habitat)" = om15)

(omodel.sel6 = modSel(oms6))

# Pick the best elevation occupancy model and then look for the effects of temperature and habitat
om16 = occu(formula = ~ obs.masl.2011 + I(obs.masl.2011^2) + obs.avg.temp.june.2011 + I(obs.avg.temp.june.2011^2) + obs.masl.2011:obs.avg.temp.june.2011 + I(obs.masl.2011^2):obs.avg.temp.june.2011 + obs.masl.2011:I(obs.avg.temp.june.2011^2) ~ obs.masl.2011 + I(obs.masl.2011^2), data = umf, knownOcc = row2011)
# "ERROR: Hessian is singular. Try providing starting values or using fewer covariates."



```



### 2012 Site occupancy modelling
Identifying the rows for the known occupancy of the 2012 detection/nondetection data
```{r}
line2012 = c()
lineobs = data.frame(Line_observations)
lineobs$Art = NULL
# for loop that appends the rows from the line observations with the year 2012 to the variable line2012
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2012) {
    line2012 = append(line2012, lineobs[row, 1])
  }
}
line2012 = matrix(line2012, ncol = 1)
line2012 = cbind(line2012, rep(2012, length(line2012)))
line2012 = cbind(line2012, rep(1, length(line2012[,1]))) # Adding 1 to all the entries of line 2012
line2012 = matrix(line2012, ncol = 3, dimnames = list(c(), c("sitenumber", "year", "number")))
line2012 = data.frame(line2012)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2012 = line2012$sitenumber
count = 0
row2012 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2012)){
    row2012 = append(row2012, count)
  }
}
```

Add detection/nondetection data and covariates together in a matrix, which makes it possible to remove the rows where there are missing covariate information. 
Overwrite the covariate variables based on the new dataset.
THE OCCU FUNCTION ONLY REMOVES ROWS WITH ALL NA's
```{r}
y2012.cov = cbind(y2012, masl.orig, precip.year.orig, avg.temp.june.orig, latitude.orig, longitude.orig, habitat.orig, sitenr)
View(y2012.cov)

y2012.cov = y2012.cov[complete.cases(y2012.cov[ , 21:25]),] # Removing rows where either of the covariates have NAs (including the y values)

y2012.cov = y2012.cov[- c(480),] # Manually removed last row because the habitat.orig had a NA value in the last row that was not caught by the complete.cases function

# Unstandardized covariates
masl = y2012.cov$masl.orig
precip = y2012.cov$precip.year.orig
avg.temp.june = y2012.cov$avg.temp.june.orig
latitude = y2012.cov$latitude.orig
longitude = y2012.cov$longitude.orig
habitat = y2012.cov$habitat.orig
latitude = y2012.cov$latitude.orig
sitenr = y2012.cov$sitenr

```

#### UnmarkedFrameOccu - 2012
Create unmarkedFrameOccu by separating the site covariates from the observation covariates
```{r}
# Create the covariates for 'obsCovs', as they need to be in a matrix form. Unstandardized!
# Habitat is regarded as the only covariate that affects the detection process
obs.habitat = matrix(data = rep(habitat, 20), nrow = 479, ncol = 20)

# Create unmarkedFrameOccu - 2012 for keeping the data
umf.12 = unmarkedFrameOccu(y = y2012.cov[1:20], siteCovs = data.frame(masl, avg.temp.june, latitude, habitat, precip), obsCovs = list(obs.habitat = obs.habitat))

```

### Model selection on the detection part - 2012
```{r}
# formula = ~ detection ~ occupancy
dmod1.2012 = occu(formula = ~ 1 ~ 1, data = umf.12, knownOcc = row2012)
summary(dmod1.2012)

dmod2.2012 = occu(formula = ~ obs.habitat ~ 1, data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
d2012.fit = fitList("p(.)psi(.)" = dmod1.2012,
               "p(habitat)psi(.)" = dmod2.2012)

(dmod.2012.sel = modSel(d2012.fit))

```

                 nPars    AIC delta   AICwt cumltvWt
p(habitat)psi(.)     6 427.92  0.00 0.99929     1.00
p(.)psi(.)           2 442.41 14.49 0.00071     1.00

The model selection shows that there is great support for the habitat affecting detection probability.

We move on to the model selection on the occupancy part with habitat included on the detection.

### Model selection on the occupancy part - 2012
```{r}
# formula = ~ detection ~ occupancy
# Looking for the effects of elevation on occupancy alone and then include habitat to see if there is a difference
om1.2012 = occu(formula = ~ 1 ~ 1, data = umf.12, knownOcc = row2012)

om2.2012 = occu(formula = ~ 1 ~ masl, data = umf.12, knownOcc = row2012)

om3.2012 = occu(formula = ~ 1 ~ masl + I(masl^2), data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit = fitList("p(.)psi(.)" = om1.2012,
              "p(.)psi(elevation)" = om2.2012,
              "p(.)psi(elevation + elevation2)" = om3.2012) # This model produces NaN's. Not enough data?
(omod.2012.sel = modSel(o2012.fit))
summary(om2.2012)
summary(om3.2012)
# The best model is the one without elevation (Finnmark causing problems again!) :)

# Looking for the effects of elevation on occupancy when habitat is included
om4.2012 = occu(formula = ~ obs.habitat ~ 1, data = umf.12, knownOcc = row2012)

om5.2012 = occu(formula = ~ obs.habitat ~ masl, data = umf.12, knownOcc = row2012)

om6.2012 = occu(formula = ~ obs.habitat ~ masl + I(masl^2), data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit1 = fitList("p(habitat)psi(.)" = om4.2012,
              "p(habitat)psi(elevation)" = om5.2012,
              "p(habitat)psi(elevation + elevation2)" = om6.2012)

(omod.2012.sel1 = modSel(o2012.fit1))

# The inclusion of habitat on the detection part as expected did not matter

# Looking for the effects of temperature on occupancy when habitat is included
om7.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june, data = umf.12, knownOcc = row2012)

om8.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june + I(avg.temp.june^2), data = umf.12, knownOcc = row2012)

om9.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june + I(avg.temp.june^2) + I(avg.temp.june^3), data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit2 = fitList("p(habitat)psi(.)" = om4.2012,
              "p(habitat)psi(temp)" = om7.2012,
              "p(habitat)psi(temp + temp2)" = om8.2012,
              "p(habitat)psi(temp+ temp2 + temp3)" = om9.2012) # NaN's produced by this model (see below)

(omod.2012.sel2 = modSel(o2012.fit2))
# Best model with psi(temp)

backTransform(linearComb(om9.2012, coefficients = c(1,0,0,0), type = 'state'))
summary(om7.2012)
summary(om8.2012)
summary(om9.2012)
# NaNs producedOccupancy (logit-scale):
#                    Estimate  SE   z  P(>|z|)
# (Intercept)          6.7156 NaN NaN     NaN
# avg.temp.june       -4.7018 NaN NaN     NaN
# I(avg.temp.june^2)   0.8511 NaN NaN     NaN
# I(avg.temp.june^3)  -0.0483 NaN NaN     NaN

# Are the NaN's produced because the dataset is so small that the optim function does not have enough data? 


# Looking for the effects of latitude on occupancy when habitat is included first, and then see it together with temperature
om10.2012 = occu(formula = ~ obs.habitat ~ latitude, data = umf.12, knownOcc = row2012)

om11.2012 = occu(formula = ~ obs.habitat ~ latitude + I(latitude^2), data = umf.12, knownOcc = row2012)

om12.2012 = occu(formula = ~ obs.habitat ~ latitude + I(latitude^2) + I(latitude^3), data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit3 = fitList("p(habitat)psi(.)" = om4.2012,
              "p(habitat)psi(lat)" = om10.2012,
              "p(habitat)psi(lat + lat2)" = om11.2012,
              "p(habitat)psi(lat + lat2 + lat3)" = om12.2012)

(omod.2012.sel3 = modSel(o2012.fit3))
# Latitude is not supported in this model

# Looking for the effects of latitude on occupancy when habitat and temp is included
om13.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june + I(avg.temp.june^2) + latitude, data = umf.12, knownOcc = row2012)

om14.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june + I(avg.temp.june^2) + latitude + I(latitude^2), data = umf.12, knownOcc = row2012)

om15.2012 = occu(formula = ~ obs.habitat ~ avg.temp.june + I(avg.temp.june^2) + latitude + I(latitude^2) + I(latitude^3), data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit4 = fitList("p(habitat)psi(temp + temp2)" = om8.2012,
              "p(habitat)psi(temp + temp2 + lat)" = om13.2012,
              "p(habitat)psi(temp + temp2 + lat + lat2)" = om14.2012, # NaN's produced
              "p(habitat)psi(temp + temp2 + lat + lat2 + lat3)" = om15.2012)
(omod.2012.sel4 = modSel(o2012.fit4))
# The summary's below show that it is only om14.2012 that produces NaN's, why is this?

backTransform(linearComb(om13.2012, coefficients = c(1,0,0,0), type = 'state'))
summary(om13.2012)
summary(om14.2012)
summary(om15.2012)

# Looking for the effects of habitat without temperature included 
om16.2012 = occu(formula = ~ obs.habitat ~ habitat, data = umf.12, knownOcc = row2012)

# Insert the fitted models into a "fitList" and rank them by AIC
o2012.fit5 = fitList("p(habitat)psi(.)"  = om4.2012,
                     "p(habitat)psi(habitat)" = om16.2012)

(omod.2012.sel5 = modSel(o2012.fit5))
# The best model is the one without habitat

# Looking for the effects of yearly precipitation without temperature included
om17.2012 = occu(formula = ~ obs.habitat ~ precip, data = umf.12, knownOcc = row2012)

om18.2012 = occu(formula = ~ obs.habitat ~ precip + I(precip^2), data = umf.12, knownOcc = row2012)

om19.2012 = occu(formula = ~ obs.habitat ~ precip + I(precip^2) + I(precip^3), data = umf.12, knownOcc = row2012)

o2012.fit6 = fitList("p(habitat)psi(.)" = om4.2012,
                     "p(habitat)psi(precip.year)" = om17.2012,
                     "p(habitat)psi(precip.year + precip.year2)" = om18.2012,
                     "p(habitat)psi(precip.year + precip.year2 + precip.year3)" = om17.2012)
(omod.2012.sel6 = modSel(o2012.fit6))
# The model with yearly precipitation psi(precip.year) included is the best

```





Identifying the rows for the known occupancy of the 2013 detection/nondetection data
```{r}
line2013 = c()

# for loop that appends the rows from the line observations with the year 2013 to the variable line2013
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2013) {
    line2013 = append(line2013, lineobs[row, 1])
  }
}
line2013 = matrix(line2013, ncol = 1)
line2013 = cbind(line2013, rep(2013, length(line2013)))
line2013 = cbind(line2013, rep(1, length(line2013[,1]))) # Adding 1 to all the entries of line 2013
line2013 = matrix(line2013, ncol = 3, dimnames = list(c(), c("sitenumber", "year", "number")))
line2013 = data.frame(line2013)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2013 = line2013$sitenumber
count = 0
row2013 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2013)){
    row2013 = append(row2013, count)
  }
}
```

Identifying the rows for the known occupancy of the 2014 detection/nondetection data
```{r}
line2014 = c()

# for loop that appends the rows from the line observations with the year 2014 to the variable line2014
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2014) {
    line2014 = append(line2014, lineobs[row, 1])
  }
}
line2014 = matrix(line2014, ncol = 1)
line2014 = cbind(line2014, rep(2014, length(line2014)))
line2014 = cbind(line2014, rep(1, length(line2014[,1]))) # Adding 1 to all the entries of line 2014
line2014 = matrix(line2014, ncol = 3, dimnames = list(c(), c("sitenumber", "year", "number")))
line2014 = data.frame(line2014)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2014 = line2014$sitenumber
count = 0
row2014 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2014)){
    row2014 = append(row2014, count)
  }
}
```

Identifying the rows for the known occupancy of the 2015 detection/nondetection data
```{r}
line2015 = c()

# for loop that appends the rows from the line observations with the year 2015 to the variable line2015
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2015) {
    line2015 = append(line2015, lineobs[row, 1])
  }
}
line2015 = matrix(line2015, ncol = 1)
line2015 = cbind(line2015, rep(2015, length(line2015)))
line2015 = cbind(line2015, rep(1, length(line2015[,1]))) # Adding 1 to all the entries of line 2015
line2015 = matrix(line2015, ncol = 3, dimnames = list(c(), c("sitenumber", "year", "number")))
line2015 = data.frame(line2015)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2015 = line2015$sitenumber
count = 0
row2015 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2015)){
    row2015 = append(row2015, count)
  }
}
```

Identifying the rows for the known occupancy of the 2016 detection/nondetection data
```{r}

line2016 = c()

# for loop that appends the rows from the line observations with the year 2016 to the variable line2016
for (row in 1:nrow(lineobs)) {
  sitenumber = lineobs[row, "RuteID"]
  year  = lineobs[row, "Aar"]
  
  if(year == 2016) {
    line2016 = append(line2016, lineobs[row, 1])
  }
}
line2016 = matrix(line2016, ncol = 1)
line2016 = cbind(line2016, rep(2016, length(line2016)))
line2016 = cbind(line2016, rep(1, length(line2016[,1]))) # Adding 1 to all the entries of line 2016
line2016 = matrix(line2016, ncol = 3, dimnames = list(c(), c("sitenumber", "year", "number")))
line2016 = data.frame(line2016)

# A for loop to identify the rows needed for the knownOccu argument in the Occu function of unmarked
sitenum2016 = line2016$sitenumber
count = 0
row2016 = c()
for(i in sitenum){
  count = count + 1
  if(any(i == sitenum2016)){
    row2016 = append(row2016, count)
  }
}
```


```{r}
y2016.cov = cbind(y2016, masl, precip.year, avg.temp.june, latitude, longitude, habitat.orig, sitenr)
View(y2016.cov)

y2016.cov = y2016.cov[complete.cases(y2016.cov[ , 21:25]),] # Removing rows where either of the covariates have NAs (including the y values)

y2016.cov = y2016.cov[- c(480),] # Manually removed last row because the habitat.orig had a NA value in the last row that was not caught by the complete.cases function

masl.2011 = y2011.cov$masl
precip.2011 = y2011.cov$precip.year
avg.temp.june.2011 = y2011.cov$avg.temp.june
latitude.2011 = y2011.cov$latitude
longitude.2011 = y2011.cov$longitude
habitat.2011 = y2011.cov$habitat.orig
latitude.2011 = y2011.cov$latitude
sitenr.2011 = y2011.cov$sitenr


```

```{r}
umf.test = unmarkedFrameOccu(y = y2016.cov[1:20], siteCovs = data.frame(avg.temp.june.2011), obsCovs = list(obs.habitat.2011 = obs.habitat.2011))

test.mod = occu(formula = ~ 1 ~ 1, data = umf.test, knownOcc = row2016)
summary(test.mod)

test.mod1 = occu(formula = ~ obs.habitat.2011 ~ 1, data = umf.test, knownOcc = row2016)

# Insert the fitted models into a "fitList" and rank them by AIC
test.fit = fitList("p(.)psi(.)" = test.mod,
               "p(habitat)psi(.)" = test.mod1)

(test.model.sel = modSel(test.fit))
```







